Master Thesis

### 2013-06-13 the-last-part.tex
Arborescence!
Writing about GACGs...

The definition of ACGs given above is quite practical in that it is
complete, it gives a precise account of what set of languages are
generated by ACGs (abstract and object languages for which there exist
the quadruples described above). This contrasts with the presentation of
ACGs that we gave in~\ref{ssec:acg}, where we defined ACGs as
collections of signatures with associated distinguished types and
connected with lexicons. Our definition only introduced the requisite
machinery, but delayed the delimitation of the way languages can be
defined. Leaving this question open will let us answer it now by
providing an extension/generalization of ACGs which will help us
overcome the modularity issues that have been the theme of this
section.\footnote{Incidentally, this was not the reason for the way we
  presented ACGs in~\ref{ssec:acg}. We chose this style as it grouped
  together the information in a more natural way. We approach signatures
  as descriptions of domains (strings, syntax, semantics), each one
  being associated with a distinguished type telling us which objects
  the domain set out to describe (strings, sentences, truth
  values). Having distinguished types directly attached to signatures
  allows us to omit them from the ACG quadruples. However, this only
  leaves us with triples containing the two signatures and the lexicon
  connecting them, and since we can infer the signatures themselves from
  the domain and the range of the lexicon, we do not have much
  motivation to introduce these tuples. Instead of composing ACGs, we
  compose just the lexicons.

On top of that, we will want to define signatures and lexicons that
yield different object languages (at least some string representation of
the form and some semantic representation of the meaning), which would
mean having at least two different ACGs which are linked (e.g. by
sharing the same abstract signature). However, since these two grammars
are very closely related, we would prefer to reason about our grammar as
\emph{one} system/graph of signatures and lexicons which defines
multiple languages instead of talking about \emph{multiple} grammars
which share some similarities.}

@@

However, a question that the authors do not pose is how to combine all
these constraints. If we were to apply the same pattern repeatedly to
get a structure like the one on (TODO: INSERT A DIAGRAM HERE), we would
have to deal with successively more complex type systems on each
level. This is due to the condition that a lexicon has to be a
homomorphism and thus it must, among other things, always map well-typed
terms to well-typed terms. To illustrate the problem, imagine a
structure such as the one on (TODO) that handles some different
constraints in each of the vertically stacked signatures including
e.g. agreement. If we tried to add a new abstract signature at the top
to implement another constraint, we would have to ensure that all the
terms we licence in our new abstract signature have a well-typed
counterpart in the existing abstract language. However, this means that
our new signature must make sure that the terms it approves do not break
agreement or any of the previously established constraints. Effectively,
our new signature would be forced to reimplement all of the existing
constraints, rendering vain all our effort to split up the constraints
into multiple signatures.

@@

We can confirm that the above graphical model of defining ACGs is
coherent. First, we can show that our interpretation is sound and so
that any directed tree adorned with signatures on the nodes, compatible
lexicons on the edges and a distinguished type on the root can only
define ACG languages. This is trivial, since the language corresponding
to the root is an ACG abstract language and the class of ACG languages
is closed under transformation by a lexicon and therefore all of the
languages corresponding to the descendants of the root are ACG languages
as well. Second, we can try and show some kind of completeness. However,
there is not much formalization done to describe a ``system of ACGs that
share common signatures'' (something that our generalization will
fix). We can nevertheless show that it is possible to express any ACG
using our graphical model. Given a grammar $\mathcal{G} =
\mathopen{<}\Sigma_A, \Sigma_O, \mathcal{L}, S\mathclose{>}$, we just
produce a tree whose root has the signature $\Sigma_A$ and distinguished
type $S$ and which is connected to a single child node having the
signature $\Sigma_O$ by the lexicon $\mathcal{L}$. Then the two nodes of
our tree define both the abstract and the object language defined by
$\mathcal{G}$.

We have now given some formal account about how we can read language
definitions in diagrams. Still, we have kept our analysis to the tree
diagrams used in current ACG literature, where every node has at most
one parent. To make the generalization more visible, we will switch from
an algebraic manner of presentation, in which we stated that the
language of a node is equal to an image of its parent's language, to a
calculus-based manner of presentation, in which we will state that the
language of a node is the set of all terms that have an antecedent in
the parent's language. Besides this property, there is also the fact
that every term in a node's language has the node's distinguished
type. This property is trivially true for nodes with parents because of
the homomorphic property of lexicons. However, it is crucial for the
root node for which this is the single defining property of its
language. To sum up, we have that for every node of a tree diagram, its
language is the set of all terms that have the node's distinguished type
and that have an antecedent in the parent's language, if there is
any. This rule is already general enough to handle the case of zero or
one parents, we can easily generalize it to handle an arbitrary count:
\emph{the language of a node is the set of all terms that have the
  node's distinguished type and that have antecedents in the languages
  of all of the node's parents}. If we transpose this back to the
algebraic style of presentation, we say that the language of a node is
the intersection of the set of terms having the node's distinguished
type and the intersection of all its parents' languages.

@@

@@

>>>>

The definition of ACGs given above is quite practical in that it is
complete, it gives a precise account of what set of languages are
generated by ACGs (abstract and object languages for which there exist
the quadruples described above). This contrasts with the presentation of
ACGs that we gave in~\ref{ssec:acg}, where we defined ACGs as
collections of signatures with associated distinguished types and
connected with lexicons. Our definition only introduced the requisite
machinery, but delayed the delimitation of the way languages can be
defined. Leaving this question open will let us answer it now by
providing an extension/generalization of ACGs which will help us
overcome the modularity issues that have been the theme of this
section.\footnote{Incidentally, this was not the reason for the way we
  presented ACGs in~\ref{ssec:acg}. We chose this style as it grouped
  together the information in a more natural way. We approach signatures
  as descriptions of domains (strings, syntax, semantics), each one
  being associated with a distinguished type telling us which objects
  the domain has set out to describe (strings, sentences, truth
  values). Having distinguished types directly attached to signatures
  allows us to omit them from the ACG quadruples. However, this only
  leaves us with triples containing the two signatures and the lexicon
  connecting them, and since we can infer the signatures themselves from
  the domain and the range of the lexicon, we do not have much
  motivation to introduce these tuples. Instead of composing ACGs, we
  compose just the lexicons.

On top of that, we will want to define signatures and lexicons that
yield different object languages (at least some string representation of
the form and some semantic representation of the meaning), which would
mean having at least two different ACGs which are linked (e.g. by
sharing the same abstract signature). However, since these two grammars
are very closely related, we would prefer to reason about our grammar as
\emph{one} system/graph of signatures and lexicons which defines
multiple languages instead of talking about \emph{multiple} grammars
which share some similarities.

TODO: Somehow express that its nice to have the links between the
signatures and lexicons that are shown in diagrams \emph{explicit}
rather than having them \emph{implicit}.}

@@

However, a question that the authors do not pose is how to combine all
these constraints. If we were to apply the same pattern repeatedly to
get a structure like the one on (TODO: INSERT A DIAGRAM HERE), we would
have to deal with successively more complex type systems on each
level. This is due to the condition that a lexicon has to be a
homomorphism and thus it must, among other things, always map well-typed
terms to well-typed terms. To illustrate the problem, imagine a
structure such as the one on (TODO) that handles some different
constraint in each of the vertically stacked signatures, including
e.g. agreement. If we tried to add a new abstract signature at the top
to implement another constraint, we would have to ensure that all the
terms we licence in our new abstract signature have a well-typed
counterpart in the existing abstract language. However, this means that
our new signature must make sure that the terms it approves do not break
agreement or any of the previously established constraints. Effectively,
our new signature would be forced to reimplement all of the existing
constraints, rendering in vain all our effort to split up the
constraints into multiple signatures.

@@

We can confirm that the above graphical model of defining ACGs is
coherent. First, we can show that our interpretation is sound and so
that any directed tree adorned with signatures on the nodes, compatible
lexicons on the edges and a distinguished type on the root can only
define ACG languages. This is trivial, since the language corresponding
to the root is an ACG abstract language and the class of ACG languages
is closed under transformation by a lexicon and therefore all of the
languages corresponding to the descendants of the root are ACG languages
as well. Second, we can try and show some kind of completeness. However,
there is not much formalization done to describe a ``system of ACGs that
share common signatures'' (something that our generalization will
fix). We can nevertheless show that it is possible to express any ACG
using our graphical model. Given a grammar $\mathcal{G} =
\mathopen{<}\Sigma_A, \Sigma_O, \mathcal{L}, S\mathclose{>}$, we just
produce a tree whose root has the signature $\Sigma_A$ and distinguished
type $S$ and which is connected to a single child node having the
signature $\Sigma_O$ by an edge labelled by the lexicon
$\mathcal{L}$. Then the two nodes of our tree define both the abstract
and the object language defined by $\mathcal{G}$.

We have now given some formal account of how we can read language
definitions in diagrams. Still, we have kept our analysis to the tree
diagrams used in current ACG literature, where every node has at most
one parent. To make the generalization more visible, we will switch from
an algebraic manner of presentation, in which we stated that the
language of a node is equal to an image of its parent's language, to a
calculus-based manner of presentation, in which we will state that the
language of a node is the set of all terms that have an antecedent in
the parent's language. Besides this property, there is also the fact
that every term in a node's language has the node's distinguished
type. This property is trivially true for nodes with parents because of
the homomorphic property of lexicons. However, it is crucial for the
root node for which this is the single defining property of its
language. To sum up, we have that for every node of a tree diagram, its
language is the set of all terms that have the node's distinguished type
and that have an antecedent in the parent's language, if there is
any. This rule is already general enough to handle the case of zero or
one parents, we can easily generalize it to handle an arbitrary count:
\emph{the language of a node is the set of all terms that have the
  node's distinguished type and that have antecedents in the languages
  of all of the node's parents}. If we transpose this back to the
algebraic style of presentation, we say that the language of a node is
the intersection of the set of terms having the node's distinguished
type and the intersection of all its parents' languages.

@@

TODO: Use the correct term arborescence in the above paragraphs!
% http://en.wikipedia.org/wiki/Arborescence_(graph_theory)

@@

\subsubsection{Graphical Abstract Categorial Grammars}

We define a \emph{graphical abstract categorial grammar} as a quadruple
$\mathcal{G} = \mathopen{<} G, \Sigma, S, {\mathcal{L}} \mathclose{>}$
where $G$ is a directed graph with vertices $V(G)$ and edges $E(G)$,
$\Sigma$ and $S$ are labelings assigning signatures and distinguished
types, respectively, to the vertices in $V$ and ${\mathcal{L}}$ is a
labeling assigning lexicons to the edges in $E$. Furthermore, a
well-formed graphical ACG satisfies the following conditions:

\begin{itemize}
\item $G$ is a directed acyclic graph.
\item For all $(u,v) \in E(G)$, $\mathcal{L}_{(u,v)} : \Sigma_u \to
  \Sigma_v$.
\item For all $(u,v) \in E(G)$, $\mathcal{L}_{(u,v)}(S_u) = S_v$.
\end{itemize}

We will say that a node $u$ is \emph{more abstract} than a node $v$
whenever $(u,v)$ belongs to the strict partial order induced by the
edges of $G$ (i.e. there is a path from $u$ to $v$).

We let the nodes in a graphical ACG $\mathcal{G} = \mathopen{<} G,
\Sigma, S, \mathcal{L} \mathclose{>}$ define two kinds of
languages. The \emph{intrinsic languages} $\mathcal{I}_{\mathcal{G}}$,
which are constrained only by the type signature described in the node
itself, and the \emph{extrinsic languages} $\mathcal{E}_{\mathcal{G}}$,
which are also constrained by type systems in more abstract nodes of the
graph. These two notions correspond to those of abstract languages and
object languages in ACGs.

$$
\mathcal{I}_{\mathcal{G}}(v) = \{t \in \Lambda(\Sigma_v)
\mid\ \vdash_{\Sigma_u} t : S_v\}
$$
$$
\mathcal{E}_{\mathcal{G}}(v) = \mathcal{I}_{\mathcal{G}}(v) \cap
\bigcap_{(u,v) \in E} \mathcal{L}_{(u,v)}(\mathcal{E}_{\mathcal{G}}(u)).
$$

From the above definition, we see that the ACG diagrams we introduced
before are a special case of graphical ACGs where the graph is an
arborescence.

\subsubsubsection{Formal Properties of Graphical ACGs}

Observation: The set of intrinsic languages definable by graphical ACGs
is the same as the set of abstract languages definable by ACGs.

Theorem: The set of extrinsic languages definable by graphical ACGs is
the set of object languages definable by ACGs closed under intersection
and transformation by a lexicon.

Proof:

First, we prove that an extrinsic language can be always constructed
from object languages by intersections and transformations by
lexicons. Let us consider any graphical ACG $\mathcal{G} = \mathopen{<}
G, \Sigma, S, \mathcal{L} \mathclose{>}$ and any topological ordering
$v_1, \ldots, v_n$ of the nodes of $G$ (an ordering such that more
abstract nodes always precede less abstract ones). Before we start, we
will just note that the all the intrinsic languages defined by $v_i$ are
abstract languages which are a special case of object languages.

\begin{itemize}
\item For $v_1$, the case is trivial since its extrinsic language is
  just its intrinsic language as the node $v_1$ has no predecessors in
  $G$.
\item For any other node $v_n$, we look at the definition of an
  extrinsic language and remark that the only operators are intersection
  and application of a lexicon with the operands being the intrinsic
  language of $v_n$ and the extrinsic languages of more abstract
  nodes. By induction hypothesis, we have that all of the operands have
  been constructed from object languages using intersections and
  transformations by lexicons and therefore even the extrinsic language
  defined by $v_n$ has this property.
\end{itemize}

Now we have to prove the converse. We do so by showing that extrinsic
languages contain object languages and are closed under intersection and
transformation by a lexicon.

\begin{itemize}
\item Let $L$ be an object language defined by the ACG $\mathopen{<}
  \Sigma_A, \Sigma_O, \mathcal{L}, S\mathclose{>}$. We take the graph
  $G$ with $V(G) = \{A, O\}$ and $E(G) = \{(A,O)\}$, we label the nodes
  $A$ and $O$ with $\Sigma_A$ and $\Sigma_O$ as signatures and $S$ and
  $\mathcal{L}(S)$ as distinguished types respectively. Finally, we
  label the edge $(A,O)$ with $\mathcal{L}$ and we have a graphical ACG
  in which the node $O$ defines the extrinsic language $L$.

\item Let $L_1$ and $L_2$ be two extrinsic languages defined by the
  nodes $u_1$ and $u_2$ respectively belonging to two graphical ACGs. We
  can construct a new graphical ACG by taking the union of the two ACGs
  (union of graphs (union of vertices and edges) and unions of the
  labelings).\footnote{Assuming, without loss of generality, that the
    sets of vertices of the two graphs are disjoint.} We then add a new
  node $v$ which will define the language $L_1 \cap L_2$.

  The signature labeling $v$ will be the union of the two signatures
  labeling $u_1$ and $u_2$ such that the intrinsic language of $v$ will
  be a superset of the intrinsic languages, and by extension the
  extrinsic languages, of both $u_1$ and $u_2$. Either of the two
  distinguished types that label $u_1$ or $u_2$ can be used to label $v$
  (if they are actually different, then the intersection is trivially
  empty). Finally, we will add two edges leading from $u_1$ and $u_2$ to
  $v$, both labeled with the identity lexicon. We will now show that
  the extrinsic language of $v$ in the newly formed graphical ACG
  $\mathcal{G}'$ is $L_1 \cap L_2$.

  \begin{align*}
    \mathcal{E}_{\mathcal{G}'}(v) &= \mathcal{I}_{\mathcal{G}'}(v) \cap
    \mathcal{L}_{(u_1,v)}(\mathcal{E}_{\mathcal{G}'}(u_1)) \cap
    \mathcal{L}_{(u_2,v)}(\mathcal{E}_{\mathcal{G}'}(u_2)) \\ &=
    \mathcal{I}_{\mathcal{G}'}(v) \cap \mathcal{E}_{\mathcal{G}'}(u_1) \cap
    \mathcal{E}_{\mathcal{G}'}(u_2) \\ &= \mathcal{E}_{\mathcal{G}'}(u_1)
    \cap \mathcal{E}_{\mathcal{G}'}(u_2) \\ &=
    \mathcal{E}_{\mathcal{G}}(u_1) \cap \mathcal{E}_{\mathcal{G}}(u_2) \\ &=
    L_1 \cap L_2
  \end{align*}

  The simplifications use the facts that: the two new lexicons are
  identities, that the intrinsic language of $v$ is a superset of the
  extrinsic languages of $u_1$ and $u_2$, that the extrinsic languages
  of $u_1$ and $u_2$ are the same in the new graphical ACG as in the two
  original graphical ACGs and finally that $L_1$ and $L_2$ are the
  extrinsic languages defined by $u_1$ and $u_2$.

\item Let $L \subset \Lambda(\Sigma_A)$ be an extrinsic language and
  $\mathcal{L} : \Sigma_A \to \Sigma_O$ a lexicon from its signature
  $\Sigma_A$ to the signature $\Sigma_O$. We will show that the language
  $\mathcal{L}(L) \subset \Lambda(\Sigma_O)$ is also an extrinsic
  language. Let $u$ be the node that defines the extrinsic language $L$
  in some graphical ACG $\mathcal{G}$. We construct a new graphical ACG
  by adding a new vertex $v$ to $\mathcal{G}$ labeled with the signature
  $\Sigma_O$ and the distinguished type $\mathcal{L}(S)$ where $S$ is
  the distinguished type labeling $u$. To this graph, we conjoin the
  edge $(u,v)$ and label it with $\mathcal{L}$. Now we verify that the
  extrinsic language of $v$ is the language $\mathcal{L}(L).$

  \begin{align*}
    \mathcal{E}_{\mathcal{G}'}(v) &= \mathcal{I}_{\mathcal{G}'}(v) \cap
    \mathcal{L}(\mathcal{E}_{\mathcal{G}'}(u)) \\
    &= \mathcal{L}(\mathcal{E}_{\mathcal{G}'}(u)) \\
    &= \mathcal{L}(\mathcal{E}_{\mathcal{G}}(u)) \\
    &= \mathcal{L}(L)
  \end{align*}

\end{itemize}

Corollary: The set of extrinsic languages definable by graphical ACGs is
the same as the set of object languages definable by ACGs if and only if
the set of object languages definable by ACGs is closed under
intersection.
